---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FeatherArray

<!-- badges: start -->
<!-- badges: end -->

`FeatherArray` makes it possible to use a [feather](https://cran.r-project.org/package=feather) file as a (read-only) [DelayedArray](https://bioconductor.org/packages/DelayedArray/) backend.

## Installation

You can install the development version of `FeatherArray` like so:

``` r
devtools::install(MalteThodberg/FeatherArray)
```

## Basic usage for matrix-like `feather` files

We load both the `feather` and `FeatherArray` packages:

```{r libs, message=FALSE}
library(feather)
library(FeatherArray)
```

Let's first simulate a large-ish matrix and write it to a feather file:

```{r example}
large_mat <- matrix(rep(rnorm(1e6), 50), ncol = 50)

feather_fname <- tempfile(fileext = ".feather")
write_feather(as.data.frame(large_mat), path = feather_fname)
```

We can then wrap the `feather` file in a `DelayedArray`. This doesn't load any data into memory yet, so it takes up very little memory:

```{r sizes}
fa <- FeatherArray(feather_fname)

object.size(fa)
object.size(large_mat)
```

All the regular read-only operations on `DelayedArray` objects are supported, e.g.:

```{r delayedarray}
fa[5:10, c(3,40, 50)]

t(fa)

fa * 4 + 1

cbind(fa, fa)
```

## Advanced usage: `feather` files with different column types

As `feather` files store data.frames, the columns can be of different types (integer, double, character, etc.). `FeatherArray` can therefore use only a subset of columns of compatible types. The default is to use all numeric types (double and integer) in the `feather` file, but this can be manually specified.

Below is an example using the iris data:

```{r iris}
iris_fname <- tempfile(fileext = ".feather")
write_feather(iris, iris_fname)

FeatherArray(iris_fname)

FeatherArray(iris_fname, columns=c("Sepal.Length", "Petal.Length"))
```

If the columns are not all of the same type, `FeatherArray` will default to sample 100 random rows and coercing them to a matrix. Alternatively users can manually specify the desired output type:

```{r type}
FeatherArray(iris_fname, type="double")
FeatherArray(iris_fname, type="integer")
FeatherArray(iris_fname, type="character")
FeatherArray(iris_fname, columns="Species", type="character")
```

## Possible extensions.

`FeatherArray` provides a basic `DelayedArray` backend for `feather` files. There are other Bioconductor functionality that could be added in the future:

- A `FeatherFile` class implemented via [BiocIO](https://bioconductor.org/packages/BiocIO/) BiocIO for basic import/export and metadata retrieval of `feather` files.
- A `FeatherDataFrame` class for random access DataFrames similar to [SQLDataFrame](https://bioconductor.org/packages/SQLDataFrame/) and [ParquetDataFrame](https://github.com/LTLA/ParquetDataFrame).
- A `SparseFeatherArray` similar to how `TENxMatrix` stores [sparse data](https://tomsing1.github.io/blog/posts/ParquetMatrix/).

The same setup used here could also be used for a `DelayedArray` backend for [parquet](https://CRAN.R-project.org/package=arrow), [fst](https://CRAN.R-project.org/package=fst) and [LaF](https://CRAN.R-project.org/package=LaF ).

